Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-05-26 cordillera-cycle.tex
Add corrections from Irina.

% Copernicus manuscript
\documentclass[tc, ms]{copernicus}

@@

% Copernicus-like latex2rtf compatible
%\input{copernicus_rtf}

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and a regional level, while little is known about the ice
sheet evolution prior to the last glacial maximum extent. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patters are available for the Laurentide and Eurasian ice sheets, such is not
the case for the Cordilleran ice sheet, where complex patterns of ice flow
emanating from multiple glaciation centres have been identified.

@@

To reconstruct the successive phases growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing
Center.

@@

The central part of an ice sheet model consists of the computation of flow
velocity which itself depends on temperature. PISM is a~shallow model, which
implies that the balance of stresses is approximated based on their predominant
components. On the other hand, the model is polythermal: it accounts for
differences in temperature and softness within the ice column.

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 31 regularly spaced layers in
bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology and was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr} and excludes variability associated to the annual
temperature cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Temperature offset time-series are reconstructed from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \aref:

@@

while temperature reconstructions from the Antarctic and oceanic cores was
provided by their authors. In addition, all records were scaled linearly in
order to simulate comparable ice volumes and realistic ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

Prior to surface mass balance computation, the model dynamically applies
a~lapse-rate correction to surface air temperature. This correction accounts
for the evolution of ice thickness on the one hand, and differences between the
climate forcing reference topography and the ice flow model basal topography on
the other hand. It uses a~reference topography distinct from the modelled basal
topography and specific to each climate forcing dataset.
An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
simulations.

@@

Despite large differences in the input temperature offset time-series, the
model output presents consistent features that can be observed across the range
of forcing data used. In all simulations, ice cover is relatively restricted
during most of the glacial cycle, while two major glaciation events occur
during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
Fig.~\ref{fig:timeseries}). A ice volume minimum is attained in between these
two events during MIS~3 (52.2--45.6\,\unit{kyr}).

On the other hand, the magitude and timing of these events vary significantly
between our simulations. For instance, GRIP and NGRIP forcing result in highest
variability in modelled ice volume, while simulations driven by oceanic records
and the NGRIP reach last glacial maximum ice volume prior to 20.0\,\unit{kyr}.

Snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). In all simulations, a central ice cap persists over
the Skeena mountains between the two glaciation events. However, the size of
this ice cap, as well as the magnitude of the MIS~4, depends on the temperature
offset time-series used to drive the model.

@@

Timing of glaciation is better reproduced by Greenland and Antarctic cores than
nearby oceanic records. This is because the oceanic cores are affected by
variations in strength of a local current (the California current) associated
with the presence of the ice sheet. The timing of the LGM is better reproduced
by Antarctic than Greenland records.

@@

During most of the glacial cycle, ice cover is restricted to high mountain
areas. This is particularly true when using the Greenland-based forcing
(Fig~\ref{fig:duration}). Most of the glacial maxima margin have a transient
character, showing that the ice sheet is far out of balance with climate. There
persist a central ice dome over the Skeena mountains over the entire glacial
period. This ice dome serves as a nucleation centre at the initiation of the
advance phase preceding the last glacial maximum. It is here because of more
widespread precipitation over the Skeena mountains than across other East-West
transects towards south or north of this region. We interpret this as an
orographic precipitation effect resulting from a "window" through the sharp
range of the Coast Mountains at this latitude (Fig?).

>>>>

% Copernicus manuscript
%\documentclass[tc, ms]{copernicus}

@@

% Copernicus-like latex2rtf compatible
\input{copernicus_rtf}

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and on regional scale, while little is known about the ice
sheet evolution prior to the last glacial maximum extent. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Laurentide and Eurasian ice sheets, such is not
the case for the Cordilleran ice sheet, where complex arrangements of ice flow
emanating from multiple glaciation centres have been identified.

@@

To reconstruct the successive phases of growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing
Center.

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 31 regularly spaced layers in
bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology and was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr} and excludes variability associated with the annual
temperature cycle itself \note{improvement} \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Temperature offset time-series are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \aref:

@@

while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors. All records were scaled linearly in
order to simulate comparable ice volumes and realistic ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

Prior to surface mass balance computation, the model dynamically applies
a~lapse-rate correction to surface air temperature. This correction accounts
for the evolution of ice thickness on the one hand, and differences between the
climate forcing reference topography and the ice flow model basal topography on
the other hand.
An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
simulations.

@@

Despite large differences in the input temperature offset time-series, the
model output presents consistent features that can be observed across the range
of forcing data used. In all simulations, ice cover is relatively restricted
during most of the glacial cycle, while two major glaciation events occur
during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
Fig.~\ref{fig:timeseries}). A local ice minimum is consistently attained in between these
two events during MIS~3 (52.2--45.6\,\unit{kyr}).

However, the magitude and timing of these events vary significantly
between our simulations. For instance, the use of GRIP and NGRIP forcing
results in highest variability in modelled ice volume. Simulations driven by
the Antarctic and GRIP records reach a last glacial maximum in ice volume at
19.5--16.9\,\unit{kyr}, while simulations driven by oceanic records
and the NGRIP record attain it prior to 20.0\,\unit{kyr}.

Despite the different timing,
snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). In all simulations, a central ice cap persists over
the Skeena mountains between the two glaciation events. However, the size of
this ice cap, as well as the magnitude of the MIS~4, depends sensitively on the temperature
offset time-series used to drive the model.

@@

Timing of glaciation is better reproduced by Greenland and Antarctic records than
nearby oceanic records. This is because the oceanic cores are affected by
variations in strength of a local current (the California current) associated
with the presence of the ice sheet. The timing of the LGM is better reproduced
by Antarctic than Greenland records.

@@

During most of the glacial cycle, ice cover is restricted to high mountain
areas. This is particularly true when using the Greenland-based forcing
(Fig~\ref{fig:duration}). Most of the glacial maxima margin has a transient
character, showing that the ice sheet is far out of balance with climate. There
persists a central ice dome over the Skeena mountains over the entire glacial
period. This ice dome serves as a nucleation centre at the initiation of the
advance phase preceding the last glacial maximum. It is here because of more
widespread precipitation over the Skeena mountains than across other East-West
transects towards south or north of this region. We interpret this as an
orographic precipitation effect resulting from a "window" through the sharp
range of the Coast Mountains at this latitude (Fig?).
